---
title: "FEAST: URT to LRT"
author: "Tejus Shinde"
date: "`r Sys.Date()`"
output: html_document
---

```{r InstallDependencies, include=FALSE, eval=FALSE}
# Run this code once before you run this analysis

# install dependencies
Packages <- c("Rcpp", "RcppArmadillo", "vegan", "dplyr", "reshape2", "gridExtra", "ggplot2", "ggthemes")
install.packages(Packages)
lapply(Packages, library, character.only = TRUE)

# install packages
devtools::install_github("cozygene/FEAST")

```

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking")

# Load the FEAST pckg
require(FEAST)
require(tidyverse)

```


## R Markdown

- **FEAST** : fast expectation-maximization for microbial source tracking - a scalable algorithm for quantifying the origins of complex microbial communities. 

This is a log for script running the tool FEAST, for the MiPORT dataset. 
Withing this project we want to run source tracking to detect URT to LRT sample microbiol transfers 

- Requires 2 mandatory **input** files:
    
    - C - An m by n count matrix, where m is the number samples and n is the number of taxa.
    - An m by 3 table, where m is the number of samples. The metadata table has three columns (i.e., ‘Env’, ‘SourceSink’, ‘id’).

<br>
- **Step 1**: Read metadata file and add factors

```{r MakeMetadata_table, cache=TRUE}

# Init vars
Data_dir="C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/"

# Read metadata file
Metadata_Df <- read.table(paste0(Data_dir,"MetaPhlan4_results/Healthy_Metaphlan4_metadata_RepeatedSamples.tsv"), header = T, sep = "\t")

```


- **Step 2**: Make a metadata file from the study metadata for FEAST

```{r preProcess_Metadata, eval=FALSE, include=FALSE}

# Add factors and filter metadata for samples of interest
    # Make a list of metadata cols to work with
    Groups_Interest <- c("SampleID", "BioProject", "Health", "SampleTypev2", "RT_category", "PatientID")

    # Filter for healthy samples first
    Metadata_Df <- Metadata_Df %>% 
      filter(Healthy == "TRUE")

    # samples retained, check counts
    table(Metadata_Df$SampleTypev2, useNA = 'ifany')

# Add factors to metadata
    # Add factors for SampleTypev2
    SamplingSite_Factor <- c("Nasal_Swab", "Buccal_mucosa", "Oral_swab", "Saliva", "Tongue_dorsum", "Supraglottal", "Sputum", "BAL", "Other")
    SamplingSite_labels <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Supraglottal", "Sputum", "BAL", "Other")
    
    Metadata_Df$SampleTypev2 <- factor(Metadata_Df$SampleTypev2, 
                                       levels = SamplingSite_Factor, 
                                       labels = SamplingSite_labels)  
    # sanity check
    # levels(Metadata_Df$SampleTypev2)

    # Add factors for RT category
    RT_Cat_Factor <- c("URT", "IRT", "LRT")
    RT_Cat_Labels <- c("Upper RT", "Intermediate RT", "Lower RT")
    Metadata_Df$RT_category <- factor(Metadata_Df$RT_category , 
                                      levels = RT_Cat_Factor,
                                      labels = RT_Cat_Labels)
    # sanity check
    # table(Metadata_Df$RT_category, useNA = 'ifany')
    # levels(Metadata_Df$RT_category)
    
  # 1. Identify sinks and sources
  sinks <- Metadata_Df %>%
    filter(RT_category == "Lower RT") %>%
    mutate(SourceSink = "Sink",
           id = row_number(),  # each sink gets a unique id
           Env = SampleTypev2)
  
  # 2. Assign same sources to each sink
  sources <- Metadata_Df %>% 
    filter(RT_category != "Lower RT") %>%
    mutate(SourceSink = "Source",
           Env = SampleTypev2) %>% 
    slice(rep(1:n(), each = nrow(sinks))) %>%
    mutate(id = rep(1:nrow(sinks), times = nrow(.) / nrow(sinks)))
  
  # 3. Combine into FEAST-compatible table
  Feast_metadata <- bind_rows(
    sinks %>% select(SampleID, Env, SourceSink, id),
    sources %>% select(SampleID, Env, SourceSink, id)
  )
  
  # 4. Preview
  head(Feast_metadata)
  
  # 5. save metadata
  write.table(Feast_metadata, "./LRT_sink_metadata.txt",
              sep = '\t', row.names = F, quote = F)
  
```

- **Step 3.A**: Read metadata table file

```{r fileReading}

# Check point: You can start here
  # Init vars
  Data_dir="C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/"

  # Add file names 
  Metadata_fileName = paste0(Data_dir,"07_Source_tracking/URT_to_LRT/LRT_sink_metadata.txt")
  Feature_fileName = paste0(Data_dir,"MetaPhlan4_results/Healthy_Metaphlan4_species_merged_RepeatedSamples_batchCorrected.tsv")

  # Read this metadata now
  Feast_metadata <- read.table(Metadata_fileName,
              sep = '\t', header = T)
  
  #Feast_metadata_sub <- Load_metadata(metadata_path = Feast_metadata)

    # sanity check
  #table(Feast_metadata$SourceSink, Feast_metadata$Env, useNA = 'ifany')
    # Apart from LRT samples (n=19, all other samples are repeated for each of the 19 LRT samples)
  print(paste("There are",length(unique(Feast_metadata$SampleID)), "number of samples in your metadata table"))



```

- **Step 3.B**: Read feature table file

```{r Read_features, echo=FALSE}

# Read feature table 
if(file.exists(Feature_fileName)){
    print("Reading abundance file...")
    Feature_df <- read.table(Feature_fileName, sep = "\t", header = T, 
                             check.names = F)
    
    # sanity check
    print(paste("Feature table has", ncol(Feature_df), "number of species &", nrow(Feature_df), "number of samples."))
    
    Feature_df <- Load_CountMatrix(CountMatrix_path = Feature_fileName)
    print(Feature_df[1:5,1:5])
    
    # Convert to pseudo-counts as FEAST expects pseudo-counts
    Feature_df_int <- ceiling(1e5 * Feature_df)

}

# Check for 0 depth samples and remove them.
# Remove samples with zero total abundance
Feature_df_int_filtered <- Feature_df_int[rowSums(Feature_df_int) > 0, ]

```


```{r FilterSamples, eval=TRUE, cache=TRUE}

# Few sanity and data format checks

zero_samples <- rownames(Feature_df_int)[rowSums(Feature_df_int) == 0]
length(zero_samples)
# View(zero_samples)

# --------------------------------- # --------------------------------- #

# If you remove samples from feature table then also remove them from here
# Create a copy of the metadata to filter
Feast_metadata_cp <- Feast_metadata

# Identify sink IDs to remove
sinkIDs_to_remove <- sapply(zero_samples, function(thisSample) {
  if (thisSample %in% Feast_metadata_cp$SampleID &&
      Metadata_Df[Metadata_Df$SampleID == thisSample, "RT_category"] == "Lower RT") {
    
    sinkID <- Feast_metadata_cp[Feast_metadata_cp$SampleID == thisSample, "id"]
    print(paste("Removing sink iteration for", sinkID))
    return(sinkID)
  } else {
    return(NA)
  }
})

# Drop NAs and make unique list of sink IDs to remove
sinkIDs_to_remove <- na.omit(unique(unlist(sinkIDs_to_remove)))

# Remove all samples with matching sink IDs
Feast_metadata <- Feast_metadata %>%
  filter(!(id %in% sinkIDs_to_remove))

# --------------------------------- # --------------------------------- #

# Check if we have all the samples in both metadata table
  #setdiff(union(row.names(Feature_df_int_filtered), unique(Feast_metadata$SampleID)),  intersect(colnames(Feature_df_int_filtered), unique(Feast_metadata$SampleID)))

  # Name check
  #stopifnot(all(row.names(Feature_df_int) %in% Feast_metadata$SampleID))

  Index <- which((row.names(Feature_df_int) %in% Feast_metadata$SampleID))
  Index <- which((Feast_metadata$SampleID %in% row.names(Feature_df_int)))
  
  #row.names(Feature_df)[Index]
  
# Check if all values are numeric
  # sapply(Feature_df, class)
  
  # if not then do this
  # Feature_df_int <- Feature_df_int %>%
  #  mutate(across(everything(), as.numeric)) 

```

- **Step 4**: Start the FEAST

Saving the output files to "./Output"


```{r enjoyThe_FEAST, echo=FALSE, cache=TRUE}

set.seed(8995)

# Specify path for output results
getwd()
OutDir="./"
setwd(OutDir)

Unique_IDs <- unique(Feast_metadata$id)

# Loop over to find samples to remove
for (eachId in Unique_IDs) {
  # Subset metadata for each sink group
  Feast_metadata_sub <- Feast_metadata %>%
    filter(id == eachId) %>% 
    mutate(id = 1) %>%  # ✅ Reset sink id to 1 for each run
    column_to_rownames("SampleID")
    
  # Define output file name
  # Set output name with the actual sample ID for sink
  sink_name <- rownames(Feast_metadata_sub[Feast_metadata_sub$SourceSink == "Sink", , drop = FALSE])[1]
  outfile_name <- paste0("LRT_sink_", sink_name)
  
  print(paste("Running FEAST on sink", eachId, "-", sink_name))

    
# run FEAST now
  FEAST_output_sub <- FEAST(C = Feature_df_int, 
                      metadata = Feast_metadata_sub, 
                      different_sources_flag = 1, 
                      dir_path = "./",
                      outfile=outfile_name,
                      EM_iterations = 1000 )
  
}



```

- **Step 5**: What the FEAST has given you?!

```{r FEASTout}

library(tidyverse)

# Check path
print(paste("You're here now:",getwd()))

# init file path
OutDir <- "C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/URT_to_LRT/Output"

# Get list of output files. Pattern starts with LRT_sink*.txt
List_Outputs <- list.files(path = OutDir, pattern = "^LRT_sink_*")

# init a df to store all results
Results_df <- data.frame()

# Read all files in the dir
for(eachReport in List_Outputs){
  
  # Read file
  fullName <- paste0(OutDir, "/",eachReport)
  
  if(file.exists(fullName)){
    print(paste("Reading output file", str_remove(eachReport, "_source_contributions_matrix.txt")))
    
    # read file
   read_df <- read.table(fullName, sep = "\t", header = T, nrows = 2, check.names = F) 
  }
  
  # rbind
  Results_df <- rbind(read_df, Results_df)

}

# Sanity check. Should have m*n matrix, where m is number of sink/rows and n is the number of sources/columns
# dim(Results_df) $ Should have 19 rows for BAL fluid (sink) and 959 columns (sources)

# save this table
write.table(Results_df, "Healthy_URT_IRT_to_LRT_FEAST_merged_output.txt", row.names = T, sep = '\t')

```

- **Step 6**: Summarize the output

```{r Summarise_contribs}

library(tidyverse)

# Check path
print(paste("You're here now:",getwd()))

# init file path
OutDir <- "C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/URT_to_LRT/Output"

# Read merged output df
Results_df <- read.table("Healthy_URT_IRT_to_LRT_FEAST_merged_output.txt", row.names = 1, sep = '\t', header = T, check.names = F)

# Melt the dataframe
Results_long <- Results_df %>%
  rownames_to_column("Sink_SampleIDs") %>%
  pivot_longer(cols = all_of(colnames(Results_df)), names_to = "Src_SampleIDs", values_to = "Value", values_drop_na = TRUE) %>%
  mutate(
    Sink_SampleID = str_extract(Sink_SampleIDs, "^[^_]+"),
    Sink_SampleType = str_extract(Sink_SampleIDs, "[^_]+$")
  ) %>%
  # Separate Src_SampleID and Src_SampleType
  mutate(
    Src_SampleID = str_extract(Src_SampleIDs, "^[^_]+"),
    Src_SampleType = str_extract(Src_SampleIDs, "[^_]+$")
  )

table(Results_long$Src_SampleType, useNA = 'ifany')

# Add factors
# Add factors for SampleType
    Src_SamplingSite_Factor <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Sputum", "Other", "Unknown")
    Src_SamplingSite_labels <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Sputum", "Other", "Unknown")
    
Results_long$Src_SampleType <- factor(Results_long$Src_SampleType,
                                      levels = Src_SamplingSite_Factor,
                                      labels = Src_SamplingSite_labels,
                                      ordered = T)

Results_long$Src_SampleType <- droplevels(Results_long$Src_SampleType)

Results_long$Sink_SampleType <- factor(Results_long$Sink_SampleType)

# Define custom colors for sampletypes + "Unknown"
SampleType_custom_colors <- c(
  "Nasal Swab" = "#E7298A",         # New, magenta-pink
  "Buccal mucosa" = "#F781BF",      # Same as original
  "Oral swab" = "#FF7F00",          # Reused from "Palatine Tonsils"
  "Saliva" = "#A65628",             # Same as original
  "Tongue dorsum" = "#4DAF4A",      # Same as original
  "Sputum" = "#E41A1C",             # Same as original
  "Other" = "#999999",              # Gray
  "Unknown" = "#CCCCCC"             # Light gray for unknown
)
  
# sanity check
    # levels(Results_long$Src_SampleType)


```

- **Step 7**: Plotting SampleType contributions for healthy

Default plot for the output

```{r def_plotSrc_contributions, eval=FALSE}
# plot cmd
PlotSourceContribution(SinkNames = rownames(Results_df)[c(1:2)],
                       SourceNames = colnames(Results_df), 
                       dir_path = OutDir,
                       mixing_proportions = Results_df, 
                       Plot_title = "URT-IRT to LRT", 
                       Same_sources_flag = 0, 
                       N = 2)


```


Boxplot for Ggplot


```{r plotSrc_contributions_bar, eval=FALSE}

# Plot 1: Box plot source contributions by Sample type
Results_long_mean <- Results_long %>% 
  group_by(Sink_SampleID, Src_SampleType) %>% 
  summarise(Proport = median(Value, na.rm=TRUE),
            Sink_SampleType = first(Sink_SampleType),
            .groups = "drop") %>%
  ungroup()

ggplotObj <- ggplot(Results_long_mean, aes(x = Sink_SampleType, y = Proport, fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_bar(stat = 'identity') +
  labs(title = "Microbial contribution from \ndifferent sample types",
       fill = "Source sample type",
       x = NULL,
       y = "Median percentage contribution") + 
  theme_bw() +
  theme(
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/LRT_sink_source_median_contributions_barplot_Sampletype.png",
       plot = savePlot,
       width = 25,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)

# spot check for each BAL samples
# Get an order of samples first
Sorted_samples <- Results_long %>% group_by(Sink_SampleID, Src_SampleType) %>% summarise(Proport = sum(Value)) %>% 
  filter(Src_SampleType %in% c("Unknown")) %>% 
  group_by(Src_SampleType) %>%
  arrange(Proport) %>% pull(Sink_SampleID)

Sorted_samples_l2 <- Results_long %>% 
  filter(!Sink_SampleID %in% Sorted_samples[14:18]) %>%
  group_by(Sink_SampleID, Src_SampleType) %>% summarise(Proport = sum(Value)) %>% 
  filter(Src_SampleType %in% c("Saliva")) %>% 
  group_by(Src_SampleType) %>%
  arrange(Proport) %>% pull(Sink_SampleID)
  
# Make a custom list
  New_list <- c(Sorted_samples_l2, Sorted_samples[14:18])
    
  # Add this list as a sampleId factor
  Results_long$Sink_SampleID <- factor(Results_long$Sink_SampleID,
                                        levels = New_list,
                                       ordered = T)


  # sanity check
  levels(Results_long$Sink_SampleID)
  
# Plot 2: Box plot source contributions by Sample type
ggplotObj <- ggplot(Results_long, aes(x = Sink_SampleID, y = Value, fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_bar(stat = 'identity') +
  labs(title = "Microbial contribution to BAL from \ndifferent sample types",
       fill = "Source sample type",
       x = NULL,
       y = "Percentage contribution") + 
  scale_fill_manual(values = SampleType_custom_colors) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/Healthy_LRT_sink_source_contributions_barplot_v2_Sampletype.png",
       plot = savePlot,
       width = 25,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)


```

- Boxplot for each sampleType


```{r plotSrc_contributions_bar, eval=FALSE}

# Plot 3: Box plot source contributions by Sample type
ggplotObj <- ggplot(Results_long, aes(x = Src_SampleType, y = (Value), fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_boxplot() +
  labs(title = "Healthy microbial contribution to LRT (BAL)\nfrom different URT sample types",
       fill = "Source sample type",
       x = NULL,
       y = "Percentage contribution + 1e-4") + 
  theme_bw() +
  theme(
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/LRT_sink_source_contributions_boxplot_Sampletype.png",
       plot = savePlot,
       width = 50,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)

```













