---
title: "FEAST: Healthy URT/LRT to IRT (Diseased samples)"
author: "Tejus Shinde"
date: "`r Sys.Date()`"
output: html_document
---

```{r InstallDependencies, include=FALSE, eval=FALSE}
# Run this code once before you run this analysis

# install dependencies
Packages <- c("Rcpp", "RcppArmadillo", "vegan", "dplyr", "reshape2", "gridExtra", "ggplot2", "ggthemes")
install.packages(Packages)
lapply(Packages, library, character.only = TRUE)

# install packages
devtools::install_github("cozygene/FEAST")

```

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Desktop path
setwd("C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/02_URT-LRT_to_IRT/")

# Laptop path
#setwd("C:/Users/o_shinde/Nextcloud/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/02_URT-LRT_to_IRT")

# Load the FEAST pckg
require(FEAST)
require(tidyverse)

```


## R Markdown

- **FEAST** : fast expectation-maximization for microbial source tracking - a scalable algorithm for quantifying the origins of complex microbial communities. 

This is a log for script running the tool FEAST, for the MiPORT dataset (diseased samples). 
Withing this project we want to run source tracking to detect URT to LRT & IRT to LRT sample microbiol transfers. 

- Requires 2 mandatory **input** files:
    
    - C - An m by n count matrix, where m is the number samples and n is the number of taxa.
    - An m by 3 table, where m is the number of samples. The metadata table has three columns (i.e., ‘Env’, ‘SourceSink’, ‘id’).

<br>
- **Step 1**: Read metadata file and add factors

```{r MakeMetadata_table, cache=TRUE, eval=FALSE}

# Read metadata file
Metadata_Df <- read.table("../../MetaPhlan4_results/Repeated_Measure_merged/All_Metaphlan4_metadata_RepeatedSamples.tsv", header = T, sep = "\t")

```


- **Step 2**: Make a metadata file from the study metadata for FEAST

```{r preProcess_Metadata, eval=FALSE, include=FALSE}

# Add factors and filter metadata for samples of interest
  # Make a list of metadata cols to work with
    Groups_Interest <- c("SampleID", "BioProject", "Health", "SampleTypev2", "RT_category", "PatientID", "Disease")

  # 374 IRT samples as sink in total for Healthy 262, CF 53, Covid 59
  # For sources, it's 19+694 = 713

    table(Metadata_Df$Disease, Metadata_Df$RT_category, useNA = "ifany")
 
  # Init disease of interest
  Disease_of_interest <- c("Healthy", "Covid-19", "Cystic Fibrosis")

# Filter Metadata for disease of interest
    table(Metadata_Df$Disease, Metadata_Df$SampleTypev2 , useNA = "ifany")
  
  # Filter samples for disease of interest: Exclude other sampletypes?
  Metadata_Df_filtered <- Metadata_Df %>%
    filter(Disease %in% Disease_of_interest)
    
  # 374 (59+53+262) IRT samples 
    table(Metadata_Df_filtered$Disease, Metadata_Df_filtered$SampleTypev2, useNA = "ifany")
    
      table(Metadata_Df_filtered$Disease, Metadata_Df_filtered$RT_category, useNA = "ifany")
  
  # IRT category and all these Disease samples are sink samples
  # All RT category and healthy samples are our sources ().
  
  '
                  IRT LRT URT
  Covid-19         59 103   0
  Cystic Fibrosis  53   0   0
  Healthy         262  19 694
  '
    # out of which Sputum (N=68), Supraglottal (59), Tongue dorsum (N = 211)
    table(Metadata_Df_filtered$RT_category, Metadata_Df_filtered$SampleTypev2 , useNA = "ifany")
    
    # Therefore, we have 68 sinks (which are all Sputum) samples
    # according to RT cat
    '
          BAL Buccal mucosa Nasal swab Oral swab Other Saliva Sputum
  IRT   0             0          0         0    36      0     68
  LRT 122             0          0         0     0      0      0
  URT   0           336         18       171     0    169      0
     
      Supraglottal Tongue dorsum
  IRT           59           211
  LRT            0             0
  URT            0             0
    '


  table(Metadata_Df_filtered$SampleTypev2, useNA = 'ifany')
  # Again, 68 samples are sinks
  '
          BAL Buccal mucosa    Nasal swab     Oral swab         Other 
          122           336            18           171            36 
       Saliva        Sputum  Supraglottal Tongue dorsum 
          169            68            59           211 
  '
  
  table(Metadata_Df_filtered$Disease, useNA = 'ifany')
  
  # And 975 Healthy - 15 Healthy sputum = 960 healthy samples are sources (19+336+18+171+36+169+211)
  
  table(Metadata_Df_filtered$Disease, Metadata_Df_filtered$SampleTypev2)
  '
                  BAL Buccal mucosa Nasal swab Oral swab Other Saliva
  Covid-19        103             0          0         0     0      0
  Cystic Fibrosis   0             0          0         0     0      0
  Healthy          19           336         18       171    36    169
                 
                  Sputum Supraglottal Tongue dorsum
  Covid-19             0           59             0
  Cystic Fibrosis     53            0             0
  Healthy             15            0           211
  '
  
  # Samples retained, check counts
    table(Metadata_Df_filtered$SampleTypev2, useNA = 'ifany')

# Add factors to metadata
    # Add factors for SampleTypev2
    SamplingSite_Factor <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Supraglottal", "Sputum", "BAL", "Other")
    SamplingSite_labels <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Supraglottal", "Sputum", "BAL", "Other")
    
    Metadata_Df_filtered$SampleTypev2 <- factor(Metadata_Df_filtered$SampleTypev2, 
                                       levels = SamplingSite_Factor, 
                                       labels = SamplingSite_labels)  
    # sanity check
    table(Metadata_Df_filtered$SampleTypev2, useNA = 'ifany')
    # levels(Metadata_Df$SampleTypev2)

    # Add factors for RT category
    table(Metadata_Df_filtered$RT_category, useNA = 'ifany')
    
    RT_Cat_Factor <- c("URT", "IRT", "LRT")
    RT_Cat_Labels <- c("Upper RT", "Intermediate RT", "Lower RT")
    Metadata_Df_filtered$RT_category <- factor(Metadata_Df_filtered$RT_category , 
                                      levels = RT_Cat_Factor,
                                      labels = RT_Cat_Labels)
    
    # sanity check
    # table(Metadata_Df$RT_category, useNA = 'ifany')
    # levels(Metadata_Df$RT_category)
    
  # 1. Identify sinks and sources
    # All Sputum (N=68) samples are sinks
  sinks <- Metadata_Df_filtered %>%
    filter(SampleTypev2 == "Sputum") %>%
    mutate(SourceSink = "Sink",
           id = row_number(),  # each sink gets a unique id
           Env = SampleTypev2)
  
  # 2. All healthy samples and URT/LRT are sources
  sources <- Metadata_Df_filtered %>% 
    filter(SampleTypev2 != "Sputum",
           Disease == "Healthy") %>%
    mutate(SourceSink = "Source",   # Add a source value column
           Env = SampleTypev2) %>%  # Add an Env value column. Usually Sample type/group  of interest
    slice(rep(1:n(), each = nrow(sinks))) %>% # Add an ID and copy that set to the number of sinks that we have. 
    mutate(id = rep(1:nrow(sinks), times = nrow(.) / nrow(sinks)))
  # slice(): index rows by their (integer) locations. 
  
  # Combine into FEAST-compatible table
  Feast_metadata <- bind_rows(
    sinks %>% select(SampleID, Env, SourceSink, id, RT_category, Disease),
    sources %>% select(SampleID, Env, SourceSink, id, RT_category, Disease)
  )
  
    
# We have 68 sputum samples as sinks and remaining 960 samples as sources. 960*68 = 65280 unique combinations
  
  # sanity check
  table(Feast_metadata$SourceSink, useNA = 'ifany')
  
  '  Sink Source 
    68  65280 '
  # Drop unused levels from Env
  table(Feast_metadata$Env, useNA = 'ifany')
  Feast_metadata$Env <- droplevels(Feast_metadata$Env)

  
  # All sinks should be Sputum samples. No Sputum samples in sources.
  table(Feast_metadata$Env, Feast_metadata$SourceSink, useNA = 'ifany')
  '
                 Sink Source
  Nasal swab        0   1224
  Buccal mucosa     0  22848
  Oral swab         0  11628
  Saliva            0  11492
  Tongue dorsum     0  14348
  Sputum           68      0
  BAL               0   1292
  Other             0   2448
  '
  
  # All sources should be healthy
  table(Feast_metadata$Disease, Feast_metadata$SourceSink, useNA = 'ifany')
  '
                   Sink Source
  Cystic Fibrosis    53      0
  Healthy            15  65280
  '
  
  # Sanity check: sample types
  table(Feast_metadata$Env, Feast_metadata$Disease, useNA = 'ifany')
  '
                  Cystic Fibrosis Healthy
  Nasal swab                  0    1224
  Buccal mucosa               0   22848
  Oral swab                   0   11628
  Saliva                      0   11492
  Tongue dorsum               0   14348
  Sputum                     53      15
  BAL                         0    1292
  Other                       0    2448
  '
  
  # Total size for metadata should be
  print(paste("#Sources-", nrow(sources), "+ #Sinks-", nrow(sinks), " = ",nrow(sinks) + nrow(sources)))
  
  # 65348
  
  # sanity check
  table(Feast_metadata$id)
  
  # 4. Preview
  head(Feast_metadata)
  
  # 5. save metadata
  write.table(Feast_metadata, "./IRT_sink_metadata_All.txt",
              sep = '\t', row.names = F, quote = F)
  
```

- **Step 3.A**: Read metadata table file

```{r fileReading}

# Check point: You can start here
  # Init vars

# for desktop
  Data_dir="C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/"

# for laptop
  Data_dir='C:/Users/o_shinde/Nextcloud/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/'
  
  # Add file names 
  Metadata_fileName = paste0(Data_dir,"07_Source_tracking/02_URT-LRT_to_IRT/IRT_sink_metadata_All.txt")
  Feature_fileName = paste0(Data_dir,"MetaPhlan4_results/Repeated_Measure_merged/All_Metaphlan4_species_merged_RepeatedSamples_batchCorrected.tsv")

  # Read this metadata now
  Feast_metadata <- read.table(Metadata_fileName,
              sep = '\t', header = T)
  
  #Feast_metadata_sub <- Load_metadata(metadata_path = Feast_metadata)

  # sanity check
  table(Feast_metadata$Env, Feast_metadata$Disease, useNA = 'ifany')
  '
                  Covid-19 Healthy Lung cancer Pneumonia
  Nasal swab           0    6210           0         0
  Buccal mucosa        0  115920           0         0
  Oral swab            0   58995           0         0
  Saliva               0   58305           0         0
  Tongue dorsum        0   72795           0         0
  Sputum               0    5175           0         0
  BAL                103      19          29       194
  Other                0   12420           0         0
  '

```

- **Step 3.B**: Read feature table file

```{r Read_features, echo=TRUE}

# Read feature table 
if(file.exists(Feature_fileName)){
    print("Reading abundance file...")
    Feature_df <- read.table(Feature_fileName, sep = "\t", header = T, 
                             check.names = F)
    
    # sanity check
    print(paste("Feature table has", ncol(Feature_df), "number of species &", nrow(Feature_df), "number of samples."))
    
    Feature_df <- Load_CountMatrix(CountMatrix_path = Feature_fileName)
    print(Feature_df[1:5,1:5])
    
    # Convert to pseudo-counts as FEAST expects pseudo-counts
    Feature_df <- ceiling(1e5 * Feature_df)

}

# Check if all values are numeric
  # sapply(Feature_df, class)
  
  # if not then do this
  # Feature_df_int <- Feature_df_int %>%
  #  mutate(across(everything(), as.numeric)) 


```


```{r FilterSamples, eval=TRUE, cache=TRUE}

# filter samples with zero total abundance
    # Check for 0 depth samples and remove them.
    # Remove samples with zero total abundance
    Feature_df_int_filtered <- Feature_df[rowSums(Feature_df) > 0, ]

# Few sanity and data format checks
  zero_samples <- rownames(Feature_df)[rowSums(Feature_df) == 0]
  
  # [1] "ERR3844798"  "ERR4791684"  "ERR4853133"  "SRR12397846"
  length(zero_samples)
  # View(zero_samples)

# Filter Feature table according to filtered metadata sample ID
  print(paste("There are",length(unique(Feast_metadata$SampleID)), "number of samples in your metadata table"))

# Matching sample IDs should be of 1301 length
  Matching_samples_Ids <- intersect(Feast_metadata$SampleID, 
                                  row.names(Feature_df_int_filtered))
  
  # Check which samples are missing
  setdiff(Feast_metadata$SampleID, row.names(Feature_df_int_filtered))
  # [1] "SRR12397846"
  # Both zero abundance samples and distinct samples are the same.
  # Remove these low abundance samples.
  
# --------------------------------- # --------------------------------- #


# If you remove samples from feature table then also remove them from FEAST metadata. 

# That means 2 things, 
  # 1. if the sample is a sink then remove the whole iteration for sink and source (specified by the id variable)
  # 2. if the sample id is not a sink then just remove the rows where it's a source

# Create a copy of the metadata to filter
Feast_metadata_cp <- Feast_metadata

# A. Remove all samples with matching sink IDs
# Identify sink IDs to remove
sinkIDs_to_remove <- sapply(zero_samples, function(thisSample) {
    if (thisSample %in% Feast_metadata$SampleID &&
        Metadata_Df_filtered[Metadata_Df_filtered$SampleID == thisSample, "SampleTypev2"] == "Sputum") {
      
      sinkID <- Feast_metadata[Feast_metadata$SampleID == thisSample, "id"]
      print(paste("Removing sink iteration for ID:", sinkID))
      return(sinkID)
    } else {
      return(NA)
    }
  })

  # Drop NAs and make unique list of sink IDs to remove
  sinkIDs_to_remove <- na.omit(unique(unlist(sinkIDs_to_remove)))

  # Remove all samples with matching sink IDs
  Feast_metadata <- Feast_metadata %>%
    filter(!(id %in% sinkIDs_to_remove))
  


# B. If it's not sink then run this to remove all samples with matching sample IDs which are not sinks
Feast_metadata <- Feast_metadata %>%
  filter(!(SampleID %in% zero_samples))

  # There should be 4 less samples in this now. So, 1027
  print(paste("There are",length(unique(Feast_metadata$SampleID)), "number of samples in your metadata table"))
  
# check if both metadata and feature tables have same sampleIDs
# Matching sample IDs should be of 1027 length
  Matching_samples_Ids <- intersect(Feast_metadata$SampleID, 
                                  row.names(Feature_df_int_filtered))

# Subset Feature_df_int_filtered for common ones. 1684800 > 1011660
Feature_df_int_filtered <- Feature_df_int_filtered[
  row.names(Feature_df_int_filtered) %in% Matching_samples_Ids,]

# --------------------------------- # --------------------------------- #

  # Final mactching check
  stopifnot(all(row.names(Feature_df_int_filtered) %in% Feast_metadata$SampleID))

  # Should be empty
  Index <- which(!(row.names(Feature_df_int_filtered) %in% Feast_metadata$SampleID))
  Index <- which(!(Feast_metadata$SampleID %in% row.names(Feature_df_int_filtered)))
  # rm if empty
  rm(Index)
  #row.names(Feature_df)[Index]
  

```

- **Step 4**: Start the FEAST

Saving the output files to "./Output"


```{r enjoyThe_FEAST, echo=FALSE, cache=TRUE}

set.seed(8995)

# Specify path for output results
getwd()
OutDir="./"
setwd(OutDir)

Unique_IDs <- unique(Feast_metadata$id)
# 68 ids retained.

# Loop over to find samples to remove
for (eachId in Unique_IDs) {
  # Subset metadata for each sink group
  Feast_metadata_sub <- Feast_metadata %>%
    filter(id == eachId) %>% 
    mutate(id = 1) %>%  # ✅ Reset sink id to 1 for each run
    column_to_rownames("SampleID")
    
  # Define output file name
  # Set output name with the actual sample ID for sink
  sink_name <- rownames(Feast_metadata_sub[Feast_metadata_sub$SourceSink == "Sink", , drop = FALSE])[1]
  disease_sink <- Feast_metadata_sub[Feast_metadata_sub$SourceSink == "Sink", "Disease"]
  
  # Add a custom output file name with Disease type of the sink sample
  outfile_name <- paste0(disease_sink, "_IRT_sink_", sink_name)
  
  print(paste("Running FEAST on sink", eachId, "-", sink_name))

# run FEAST now
  FEAST_output_sub <- FEAST(C = Feature_df_int_filtered, 
                      metadata = Feast_metadata_sub, 
                      different_sources_flag = 1, 
                      dir_path = "./",
                      outfile=outfile_name,
                      EM_iterations = 1000 )
  
}



```

- **Step 5**: What the FEAST has given you?!

```{r FEASTout, eval=FALSE}

library(tidyverse)

# Check path
print(paste("You're here now:",getwd()))

# init file path
OutDir <- "C:/Users/o_shinde/Nextcloud/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/02_URT-LRT_to_IRT/Disease_Output/"

# Desktop init path
OutDir = "C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/02_URT-LRT_to_IRT/Diseased_Group_Output/"

# For each diseases we generate a pattern and identify files to merge them.
Disease_of_Interest
  # "Healthy"     "Cystic Fibrosis"   "Covid-19"    

# init a df to store all results
Healthy_Results_df <- data.frame()

# Get list of output files. Pattern starts with IRT_sink*.txt
List_Outputs <- list.files(path = OutDir, pattern = "*_IRT_sink_*")

# Read all files in the dir
for(eachReport in List_Outputs){
  
  # Read file
  fullName <- paste0(OutDir, "/",eachReport)
  
  if(file.exists(fullName)){
    print(paste("Reading output file", str_remove(eachReport, "_source_contributions_matrix.txt")))
    
    # read file
   read_df <- read.table(fullName, sep = "\t", header = T, nrows = 2, check.names = F) 
  }
  
  # rbind
  Healthy_Results_df <- rbind(read_df, Healthy_Results_df)

}

# Sanity check. Should have m*n matrix, where m is number of sink/rows and n is the number of sources/columns
# dim(Healthy_Results_df) $ Should have 19 rows for BAL fluid (sink) and 959 columns (sources)

# save this table
write.table(Healthy_Results_df, "Disease_Group_IRT_Sinks_from_URT_LRT_FEAST_merged_output.txt", row.names = T, sep = '\t')

```

- **Step 6**: Summarize the output

```{r Summarise_contribs, eval=FALSE}

library(tidyverse)

# Check path
print(paste("You're here now:",getwd()))

# init file path
OutDir <- "C:/Users/o_shinde/Desktop/MiPORT/AG Moissl-Eichinger_people/Tejus/MiPoRT Paper/MiPORT_DA/07_Source_tracking/02_URT-LRT_to_IRT/Plots/"

# Read merged output df
Results_df <- read.table("./Disease_Group_IRT_Sinks_from_URT_LRT_FEAST_merged_output.txt", row.names = 1, sep = '\t', header = T, check.names = F)

# Check if all row Sums add to one
RowSums_Proportions_Index <- rowSums(Results_df)
RowSums_Proportions_Boolean <- all(rowSums(Results_df) == 1)


View(Results_df[RowSums_Proportions_Index,])

# Melt the dataframe
Results_long <- Results_df %>%
  rownames_to_column("Sink_SampleIDs") %>%
  pivot_longer(cols = all_of(colnames(Results_df)), names_to = "Src_SampleIDs", values_to = "Value", values_drop_na = TRUE) %>%
  mutate(
    Temp = str_replace(Sink_SampleIDs, "BAL_BAL$", ""),
    Sink_SampleID = str_extract(Temp, "^[^_]+"),
    Sink_SampleType = str_extract(Sink_SampleIDs, "[^_]+$")
  ) %>% select(-Temp) %>%
  # Separate Src_SampleID and Src_SampleType
  mutate(
    Src_SampleID = str_extract(Src_SampleIDs, "^[^_]+"),
    Src_SampleType = str_extract(Src_SampleIDs, "[^_]+$")
  )

table(Results_long$Src_SampleType, useNA = 'ifany')/341
table(Results_long$Sink_SampleType, useNA = 'ifany')/341

# Add factors
# Add factors for SampleType
    Src_SamplingSite_Factor <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Supraglottal", "Sputum", "Other", "Unknown")
    Src_SamplingSite_labels <- c("Nasal swab", "Buccal mucosa", "Oral swab", "Saliva", "Tongue dorsum", "Supraglottal", "Sputum", "Other", "Unknown")
    
Results_long$Src_SampleType <- factor(Results_long$Src_SampleType,
                                      levels = Src_SamplingSite_Factor,
                                      labels = Src_SamplingSite_labels,
                                      ordered = T)

table(Results_long$Src_SampleType, useNA = 'ifany')/341

#Results_long$Src_SampleType <- droplevels(Results_long$Src_SampleType)

# Add factor to sink sampleType
Results_long$Sink_SampleType <- factor(Results_long$Sink_SampleType)

# Define custom colors for sampletypes + "Unknown"
SampleType_custom_colors <- c(
  "Nasal Swab" = "#E7298A",         # New, magenta-pink
  "Buccal mucosa" = "#F781BF",      # Same as original
  "Oral swab" = "#FF7F00",          # Reused from "Palatine Tonsils"
  "Saliva" = "#A65628",             # Same as original
  "Tongue dorsum" = "#4DAF4A",      # Same as original 
  "Supraglottal" = "#984EA3",       # Same as original
  "Sputum" = "#E41A1C",             # Same as original
  "Other" = "#999999",              # Gray
  "Unknown" = "#CCCCCC"             # Light gray for unknown
)
  
# sanity check
    # levels(Results_long$Src_SampleType)


```

- **Step 7**: Plotting SampleType contributions for healthy

Default plot for the output

```{r def_Disease_plotSrc_contributions, eval=FALSE}
# plot cmd
PlotSourceContribution(SinkNames = rownames(Results_df)[c(1:2)],
                       SourceNames = colnames(Results_df)[c(1:10)], 
                       dir_path = OutDir,
                       mixing_proportions = Results_df, 
                       Plot_title = "URT-IRT to LRT", 
                       Same_sources_flag = 0, 
                       N = 2)

# Doesn't make sense for samples > 10

```


Boxplot for Ggplot


```{r Disease_plotSrc_contributions_bar, eval=FALSE}

# Read metadata and join it to the long table
Feast_metadata <- read.table("IRT_sink_metadata_All.txt", header = T, 
                             sep = "\t")

# Format the sampleIDs to match
Feast_metadata_cp <- Feast_metadata

Feast_metadata_sub <- Feast_metadata %>%
  filter(SourceSink == 'Source') %>%
  mutate(SampleID = str_extract(SampleID, "^[^_]+"))

length(intersect(Results_long$Src_SampleID, Feast_metadata_sub$SampleID))

# Plot 1: Box plot source contributions by Sample type
Results_long_mean <- Results_long %>% 
  left_join(Feast_metadata, by = join_by("Src_SampleID" == "SampleID")) %>%
  group_by(Sink_SampleID, Src_SampleType) %>% 
  summarise(Proport = median(Value, na.rm=TRUE),
            Sink_SampleType = first(Sink_SampleType),
            .groups = "drop") %>%
  ungroup()

ggplotObj <- ggplot(Results_long_mean, aes(x = Sink_SampleType, y = Proport, fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_bar(stat = 'identity') +
  labs(title = "Microbial contribution from \ndifferent sample types (diseased condition)",
       fill = "Source sample type",
       x = NULL,
       y = "Median percentage contribution") + 
  theme_bw() +
  theme(
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/IRT_sink_source_median_contributions_barplot_Sampletype_Diseased.png",
       plot = savePlot,
       width = 25,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)

# spot check for each BAL samples

unknown_sorted <- Results_long %>%
  group_by(Src_SampleType) %>%
  summarise(Proport = sum(Value), .groups = "drop")

# Buccal mucosa and Sputum have the highest sum proportions after Unknown

# Step 1: Sort all samples by Unknown contribution
unknown_sorted <- Results_long %>%
  group_by(Sink_SampleID, Src_SampleType) %>%
  summarise(Proport = sum(Value)) %>%
  filter(Src_SampleType == "Unknown") %>%
  arrange(Proport) %>%
  pull(Sink_SampleID)

# Step 2: Select top 150 low-unknown samples
top150_low_unknown <- unknown_sorted[1:150]

# Step 3: Sub-sort these 150 by Sputum contribution (descending)
top150_sorted <- Results_long %>%
  filter(Sink_SampleID %in% top150_low_unknown & Src_SampleType == "Sputum") %>%
  group_by(Sink_SampleID) %>%
  summarise(SputumProp = sum(Value), .groups = "drop") %>%
  arrange(desc(SputumProp), .by_group = TRUE) %>%
  pull(Sink_SampleID)

# Step 4: Concatenate with the rest of the samples (including high Unknown)
final_sample_order <- c(top150_sorted, unknown_sorted[151:350])

  # Add this list as a sampleId factor
  Results_long$Sink_SampleID <- factor(Results_long$Sink_SampleID,
                                        levels = unknown_sorted,
                                       ordered = T)
  # rm factor
  Results_long$Sink_SampleID <- factor(Results_long$Sink_SampleID)
  
  # sanity check
  levels(Results_long$Sink_SampleID)
  
# Plot 2: Bar plot source contributions by Sample type
ggplotObj <- ggplot(Results_long, aes(x = Sink_SampleID, y = Value, fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_bar(stat = 'identity') +
  labs(title = "Microbial contribution to BAL from \ndifferent sample types (Diseased)",
       fill = "Source sample type",
       x = NULL,
       y = "Percentage contribution") + 
  ylim(0,1) + 
  scale_fill_manual(values = SampleType_custom_colors) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/Diseased_IRT_sink_source_contributions_barplot_v2_Sampletype.png",
       plot = savePlot,
       width = 25,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)


```

- Boxplot for each sampleType


```{r Disease_plotSrc_box, eval=FALSE}

# Plot 3: Box plot source contributions by Sample type
ggplotObj <- ggplot(Results_long, aes(x = Src_SampleType, y = (Value), fill = Src_SampleType))

# save plot
savePlot <- ggplotObj + geom_boxplot() +
  labs(title = "Healthy microbial contribution to IRT (Sputum)\nfrom different URT-LRT sample types",
       fill = "Source sample type",
       x = NULL,
       y = "Percentage contribution + 1e-4") + 
  theme_bw() +
  theme(
        axis.text = element_text(size = 22), 
        axis.title = element_text(size = 22, face = "bold"),
        title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20, face = "bold"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt"),
        strip.text = element_text(size = 18, face = "bold")) 

ggsave("./Plots/IRT_sink_source_contributions_boxplot_Sampletype.png",
       plot = savePlot,
       width = 50,          
       height = 18,
       units = "cm",
       dpi = 300,               # High resolution
       limitsize = FALSE)

```













